sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2023-09-12 doc/paper-jss/paper-jss.Rmd
Copy editing

$$
\begin{aligned}
\mu_{\bm{s},t} &=
f^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
O_{\bm{s},t} +
\alpha_g +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} + 
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \bm{\gamma_t} +
\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} \right),
\end{aligned}
$$
This process component is combined with an observation error family (e.g., Gaussian, Gamma, binomial, Tweedie) and link function $f^{-1}$ (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
Some families can be combined into a two-part "delta" or "hurdle" model [@aitchison1955] to model the zero vs. non-zero observations separately from the positive observations.
See the model description vignette for a full description of the statistical model (`browseVignettes("sdmTMB")` or see the \href{https://pbs-assess.github.io/sdmTMB/articles/}{rendered version}).

The GLMMs underpinning \pkg{sdmTMB} models are spatially explicit---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions (e.g., \pkg{randomForest} [@liaw2002], \pkg{MaxEnt} [@phillips2006], and most smooths in \pkg{mgcv} [@wood2017a]).
The random fields in \pkg{sdmTMB} are structured as MVN constrained by a Matérn covariance function [@matern1960].
The Matérn can accommodate a range of shapes and can be both isotropic (correlation decays the same in all directions) or anisotropic (correlation is directionally dependent) [@haskard2007].
The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to $\sim 0.13$ [@lindgren2015]---can be shared or estimated separately (`share_range` argument).

By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Figure \ref{fig:diagram}).
Turning off both spatial and spatiotemporal effects allows for comparison with standard non-spatial GLMs or GLMMs.
We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in \pkg{mgcv} [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes: hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()`.
\pkg{sdmTMB} can use many `mgcv::s()` and `mgcv::t2()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers (`s(x, bs = "cc")`), and smoothers with specified basis dimensions (`s(x, k = 4)`) [@wood2017a].

@@

The \pkg{sdmTMB} package is constructed from S3 classes and designed to be both modular and familiar to users of widely used \proglang{R} packages (e.g. \pkg{glmmTMB}, \pkg{lme4}, \pkg{mgcv}). Data may be simulated with or without a fitted model (`simulate.sdmTMB()`, `sdmTMB_simulate()` respectively). Estimation is done via maximum marginal likelihood with the `sdmTMB()` function (or `sdmTMB_cv()` for cross-validation), and the resulting fitted model can be used to make predictions (`predict.sdmTMB()`) or calculate residuals (`residuals.sdmTMB()`), derive additional quantities of interest, and simulate Bayesian posterior distributions. Common quantities of interest may include summarizing shifts in the center of gravity (`get_cog()`) [e.g., @thorson2016cog] or in the field of ecology, tracking total population densities through time (`get_index()`).

@@

```{r dog-binomial, results='hide', message=FALSE, warning=FALSE, echo=TRUE}
mesh_pcod <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 8)
plot(mesh_pcod)
```

@@

We can test for spatial autocorrelation with a Moran's I test or with a visual
inspection of the residuals calculated from the `residuals()` S3 method. We
omit the Moran's I code for brevity but include it in the code supplement. The
p-value for spatial autocorrelation is low for the model without a random field
(p < 0.01) but large for the model with a random field (p = `r p_rf`)
suggesting significant residual correlation in the simpler model that is
alleviated by including the random field.

@@

We can also use `tidy()` to get a data frame of our random field parameters,
where `sigma_O` is $\sigma_O$ in the above equations: the standard deviation of
the random field.

@@

$$
\begin{aligned}
\mathbb{E}[y_{\bm{s},t}] &= \mu_{\bm{s},t},\\
\mu_{\bm{s},t} &=
f^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
O_{\bm{s},t} +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \gamma_t +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} 
\right),
\end{aligned}
$$

@@

We can explore four alternative families that may better represent the data.
Each alternative family uses a delta or hurdle distribution: delta-gamma and
delta-lognormal (REF), and delta-Poisson-link-gamma and
delta-Poisson-link-lognormal (REF). Note that we choose to omit the
spatiotemporal random fields from the first model component 
(`spatiotemporal = list("off", "iid")`) because initial fits indicated this random field was
collapsing to zero indicating minimal changes in encounter spatial distribution
from year to year.

@@

$$
\begin{aligned}
\bm{\delta}_{t=1} &\sim \operatorname{MVNormal} (\bm{0}, \bm{\Sigma}_{\epsilon}),\\
\bm{\delta}_{t>1} &= \rho \bm{\delta}_{t-1} + \sqrt{1 - \rho^2} \bm{\epsilon_{t}},  \:
\bm{\epsilon_{t}} \sim \operatorname{MVNormal} \left(\bm{0}, \bm{\Sigma}_{\epsilon} \right),
\end{aligned}
$$

@@

Our prediction data frame is similar to the binomial spatial model, but
includes columns for the two delta model components (labelled with suffixes `1`
and `2`) and adds an `epsilon_st` column for spatiotemporal random effects. We can
easily generate plots from this data frame with ggplot2 or similar. We suppress
that code in this paper for brevity.

@@

We can generate an area-weighted population index (e.g., a relative or absolute
index of abundance or biomass) that is independent of sampling locations by
predicting from the model on a grid covering the area of interest and summing
the predicted biomass with the `get_index()` function (Figure
\ref{fig:dog-index-plot}). We supply the grid cell area (4 km^2^) to the `area`
argument and specify `bias_correct = TRUE` to enable a generic epsilon-bias
correction needed due to the non-linear transformation of the random effects
(REF).

@@

$$
\begin{aligned}
\mathbb{E}[y_{\bm{s},t}] &= \mu_{\bm{s},t},\\
\mu_{\bm{s},t} &=
f^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
\alpha_g +
\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} \right),
\end{aligned}
$$

>>>>

$$
\begin{aligned}
\mu_{\bm{s},t} &=
g^{-1} \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
O_{\bm{s},t} +
\alpha_g +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} + 
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \bm{\gamma_t} +
\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} \right),
\end{aligned}
$$
This process component is combined with an observation error family (e.g., Gaussian, gamma, binomial, Tweedie) and link function $g$ (e.g., identity, log, logit) as in any generalized linear model (GLM) [@mccullagh1989].
Some families can be combined into a two-part "delta" or "hurdle" model [@aitchison1955] to model the zero vs. non-zero observations separately from the positive observations.
See the model description vignette for an expanded description of the statistical model (`browseVignettes("sdmTMB")` or see the \href{https://pbs-assess.github.io/sdmTMB/articles/}{rendered version}).

The GLMMs underpinning \pkg{sdmTMB} models are spatially explicit---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions (e.g., \pkg{randomForest} [@liaw2002], \pkg{MaxEnt} [@phillips2006], and most smooths in \pkg{mgcv} [@wood2017a]).
The random fields in \pkg{sdmTMB} are structured as MVN constrained by a Matérn covariance function [@matern1960].
The Matérn can accommodate a range of shapes and can be both isotropic (correlation decays the same in all directions) or anisotropic (correlation is directionally dependent) [@haskard2007].
The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to $\sim 0.13$ [@lindgren2015]---can be shared among random fields or estimated separately (`share_range` argument).

By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Figure \ref{fig:diagram}).
Turning off both spatial and spatiotemporal effects allows for comparison with standard non-spatial GLMs or GLMMs.
We include additional flexibility in specifying the linear main effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in \pkg{mgcv} [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes: hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()`.
\pkg{sdmTMB} can use many `mgcv::s()` and `mgcv::t2()` smoothers including bivarate smoothers `s(x, y)`, smoothers varying by continuous or categorical variables `s(x1, by = x2)`, cyclical smoothers `s(x, bs = "cc")`, and smoothers with specified basis dimensions `s(x, k = 4)` [@wood2017a].

@@

The \pkg{sdmTMB} package is constructed from S3 classes and designed to be both modular and familiar to users of widely used \proglang{R} packages (e.g. \pkg{glmmTMB}, \pkg{lme4}, \pkg{mgcv}). Data may be simulated with or without a fitted model (`simulate.sdmTMB()`, `sdmTMB_simulate()` respectively). Estimation is done via maximum marginal likelihood with the `sdmTMB()` function, and the resulting fitted model can be used to make predictions (`predict.sdmTMB()`) or calculate residuals (`residuals.sdmTMB()`), derive additional quantities of interest, and simulate Bayesian posterior distributions. Common quantities of interest may include summarizing shifts in the center of gravity (`get_cog()`) [e.g., @thorson2016cog] or in the field of ecology, tracking total population densities through time (`get_index()`).
<!-- TODO: describe this in more detail and how it could apply beyond ecology; maybe a derived variables subsection? -->

@@

```{r dog-binomial, results='hide', message=FALSE, warning=FALSE, echo=TRUE, fig.cap="TODO"}
mesh_pcod <- make_mesh(pcod, xy_cols = c("X", "Y"), cutoff = 8)
plot(mesh_pcod)
```

@@

We can test for spatial autocorrelation with a Moran's I test or with a visual
inspection of the residuals calculated from the `residuals()` S3 method. We
omit code for Moran's I calculation for brevity but include an example in the
code supplement. The p-value for spatial autocorrelation is low for the model
without a random field (p < 0.01) but large for the model with a random field
(p = `r p_rf`) suggesting significant residual correlation in the simpler model
that is alleviated by including the random field.

@@

We can also use `tidy()` to get a data frame of our random field parameters,
where `sigma_O` is $\sigma_O$ in the above equations: the standard deviation of
the spatial random field $\bm{\omega}$.

@@

$$
\begin{aligned}
\mathbb{E}[y_{\bm{s},t}] &= \mu_{\bm{s},t},\\
\mu_{\bm{s},t} &=
\exp \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
O_{\bm{s},t} +
\bm{X}^{\mathrm{tvc}}_{\bm{s},t} \gamma_t +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} 
\right),
\end{aligned}
$$

@@

We can explore four alternative families that may better represent the data.
Each alternative family uses a delta or hurdle distribution: delta-gamma and
delta-lognormal (REF), and delta-Poisson-link-gamma and
delta-Poisson-link-lognormal (REF). Note that we choose to omit the
spatiotemporal random fields from the binomial model component 
(`spatiotemporal = list("off", "iid")`) because initial fits indicated this random field was
collapsing to zero indicating minimal changes in encounter spatial distribution
from year to year.

@@

$$
\begin{aligned}
\bm{\delta}_{t=1} &\sim \operatorname{MVN} (\bm{0}, \bm{\Sigma}_{\epsilon}),\\
\bm{\delta}_{t>1} &= \rho \bm{\delta}_{t-1} + \sqrt{1 - \rho^2} \bm{\epsilon_{t}},  \:
\bm{\epsilon_{t}} \sim \operatorname{MVN} \left(\bm{0}, \bm{\Sigma}_{\epsilon} \right),
\end{aligned}
$$

@@

Our prediction data frame is similar to the binomial spatial model, but includes
columns for the two delta model components (labelled with suffixes `1` and `2`)
and adds an `epsilon_st` column for spatiotemporal random effects. We can easily
generate plots from this data frame (e.g., with `ggplot2::geom_raster()`). We
suppress that code in this paper for brevity.

@@

We can generate an area-weighted population index (e.g., a relative or absolute
index of abundance or biomass) that is independent of sampling locations by
predicting from the model on a grid covering the area of interest and summing
the predicted biomass with the `get_index()` function (Figure
\ref{fig:dog-index-plot}). We supply the grid cell area (4 km^2^) to the `area`
argument and specify `bias_correct = TRUE` to enable a generic epsilon-bias
correction needed due to the non-linear transformation of the random effects
(REF).
<!-- TODO: and the Laplace? -->

@@

$$
\begin{aligned}
\mathbb{E}[y_{\bm{s},t}] &= \mu_{\bm{s},t},\\
\mu_{\bm{s},t} &=
\exp \left( \bm{X}^{\mathrm{main}}_{\bm{s},t} \bm{\beta} +
\alpha_g +
\bm{X}^{\mathrm{svc}}_{\bm{s},t} \zeta_{\bm{s}} +
\omega_{\bm{s}} +
\epsilon_{\bm{s},t} \right),
\end{aligned}
$$

where TODO.
TODO: get NB2 above and similar in previous examples
